Interpreting Random Forest Predictions for Firearm Identification Using LIME

### 2018-12-04 paper/Interpreting_Random_Forest_Predictions_for_Firearm_Identification_Using_LIME.Rmd
a bunch of work from the last week that I forgot to commit - removed subsampling and reran code - cleaned up app code and improved app - started work on assessing variability

# Load libraries
library(tidyverse)
library(bulletr)
library(randomForest)
library(lime)
library(cowplot)
library(purrr)

@@

```{r}
# Create a subsampled version of the training data
if (!file.exists("../data/hamby173and252_train_sub.csv")) {

  # Create the data (with a set seed)
  set.seed(20181128)
  hamby173and252_train_sub <- hamby173and252_train %>%
    filter(samesource == FALSE) %>%
    slice(sample(1:length(hamby173and252_train$barrel1), 4500, replace = FALSE)) %>%
    bind_rows(hamby173and252_train %>% filter(samesource == TRUE))

  # Save the subsampled data
  write.csv(hamby173and252_train_sub, "../data/hamby173and252_train_sub.csv", row.names = FALSE)

  # Load in the subsampled training data
  hamby173and252_train_sub <- read.csv("../data/hamby173and252_train_sub.csv")

@@

```{r fig.height = 5, fig.width = 8}
# Create plots of the feature distributions colored by samesource for the training data
hamby173and252_train %>% 
  select(rf_features, samesource) %>%
  gather(key = feature, value = value, 1:9) %>%
  select(feature, value, samesource) %>%
  ggplot(aes(x = value, fill = samesource)) + 
  geom_histogram(bins = 30) + 
  facet_wrap( ~ feature, scales = "free") +
  labs(x = "Variable Value", y = "Frequency", fill = "Same Source?") +
  theme_bw() +
  theme(legend.position = "bottom")

# Same plot but with the subsampled training data
hamby173and252_train_sub %>% 
  select(rf_features, samesource) %>%
  gather(key = feature, value = value, 1:9) %>%
  select(feature, value, samesource) %>%
  ggplot(aes(x = value, fill = samesource)) + 
  geom_histogram(bins = 30) + 
  facet_wrap( ~ feature, scales = "free") +
  labs(x = "Variable Value", y = "Frequency", fill = "Same Source?") +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r fig.height = 5, fig.width = 8}
# Create plots of the feature distributions colored by samesource for the training data
# using the position = "fill" option
hamby173and252_train %>% 
  select(rf_features, samesource) %>%
  gather(key = feature, value = value, 1:9) %>%
  select(feature, value, samesource) %>%
  ggplot(aes(x = value, fill = samesource)) + 
  geom_histogram(position = "fill", bins = 30) + 
  facet_wrap( ~ feature, scales = "free") +
  labs(x = "Variable Value", y = "Proportion", fill = "Same Source?") +
  theme_bw() +
  theme(legend.position = "bottom")

# Same plot but with the subsampled training data
hamby173and252_train_sub %>% 
  select(rf_features, samesource) %>%
  gather(key = feature, value = value, 1:9) %>%
  select(feature, value, samesource) %>%
  ggplot(aes(x = value, fill = samesource)) + 
  geom_histogram(position = "fill", bins = 30) + 
  facet_wrap( ~ feature, scales = "free") +
  labs(x = "Variable Value", y = "Proportion", fill = "Same Source?") +
  theme_bw() +
  theme(legend.position = "bottom")
```

@@

  # Specify the options to use with lime
  hamby224_lime_inputs <- data.frame(data_name = c(rep("Full", 12), rep("Subsampled", 12)),
                                     bin_continuous = rep(c(rep(TRUE, 10), rep(FALSE, 2)), 2),
                                     quantile_bins = rep(c(rep(TRUE, 5), rep(FALSE, 5), rep(TRUE, 2)), 2),
                                     nbins = rep(c(rep(2:6, 2), rep(4, 2)), 2),
                                     use_density = rep(c(rep(TRUE, 10), TRUE, FALSE)), 2) %>%
    mutate(case = 1:length(data_name)) %>%
    select(case, data_name, bin_continuous, nbins, quantile_bins, use_density)

  # Apply lime to the full training data with the specified input options
  hamby224_lime_explain_full <- pmap(.l = as.list(hamby224_lime_inputs %>% 
                                                    filter(data_name == "Full") %>%
                                                    select(-case, -data_name)),
             .f = run_lime, # run_lime is one of my helper functions
             data_name = "Full",
             train = hamby173and252_train %>% select(rf_features),
             test = hamby224_test %>% arrange(case) %>% select(rf_features) %>% na.omit(),
             rfmodel = as_classifier(rtrees),
             label = "TRUE",
             nfeatures = 3,
             seed = TRUE)

  # Separate the lime and explain function results from the full data
  hamby224_lime_full <- map(hamby224_lime_explain_full, function(list) list$lime)
  hamby224_explain_full <- map_df(hamby224_lime_explain_full, function(list) list$explain)

  # Apply lime to the subsampled training data with the specified input options
  hamby224_lime_explain_sub <- pmap(.l = as.list(hamby224_lime_inputs %>% 
                                               filter(data_name == "Subsampled") %>%
                                               select(-case, -data_name)),
             .f = run_lime, # run_lime is one of my helper functions
             data_name = "Subsampled",
             train = hamby173and252_train_sub %>% select(rf_features),
             test = hamby224_test %>% arrange(case) %>% select(rf_features) %>% na.omit(),
             rfmodel = as_classifier(rtrees),
             label = "TRUE",
             nfeatures = 3,
             seed = TRUE)

  # Separate the lime and explain function results from the subsampled data
  hamby224_lime_sub <- map(hamby224_lime_explain_sub, function(list) list$lime)
  hamby224_explain_sub <- map_df(hamby224_lime_explain_sub, function(list) list$explain)

  # Join the lime results from the full and subsampled training data
  hamby224_lime <- c(hamby224_lime_full, hamby224_lime_sub)
  hamby224_explain <- bind_rows(hamby224_explain_full, hamby224_explain_sub)

  # Name the items in the lime list
  names(hamby224_lime) <- map_chr(1:dim(hamby224_lime_inputs)[1], function(case) 
      sprintf("case: data = %s, bin_continuous = %s, quantile_bins = %s, nbins = %0.f, use_density = %s",
              hamby224_lime_inputs$data_name[case],
              hamby224_lime_inputs$bin_continuous[case],
              hamby224_lime_inputs$quantile_bins[case],
              hamby224_lime_inputs$nbins[case],
              hamby224_lime_inputs$use_density[case]))

@@

  # Join the data and the explanations and edit and add additional variables
  # Create the feature bin labels using my function "bin_labeller"
  hamby224_test_explain <- hamby224_test %>%
    mutate(case = as.character(case)) %>%
    full_join(hamby224_explain, by = "case") %>%
    mutate(case = factor(case),
           feature_desc = factor(feature_desc),
           feature_bin = pmap_chr(list(feature = feature, 
                                  feature_value = feature_value,
                                  d_n = training_data, 
                                  b_c = bin_continuous,
                                  q_b = quantile_bins,
                                  n_b = nbins,
                                  u_d = use_density),
                            .f = bin_labeller, # bin_labeller is one of my helper functions
                            bin_data = hamby224_bin_boundaries,
                            case_info = hamby224_lime_inputs)) %>%
    mutate(feature = factor(feature),
           nbins = factor(nbins),
           feature_number = readr::parse_number(feature_desc),
           strictly_less = FALSE) %>%
    arrange(nbins)

@@

  # Reorder the variables of feature_desc and feature_bin for plotting purposes
  hamby224_test_explain <- hamby224_test_explain %>%
    mutate(feature_desc = reorder(feature_desc, strictly_less),
           feature_desc = reorder(feature_desc, feature_number),
           feature_desc = reorder(feature_desc, as.numeric(feature))) %>%
    select(training_data:use_density, case:feature_desc, feature_bin:strictly_less, 
           data, prediction)

@@

  # Create a data frame with the interesting information relating to the differe
  # evaluations of lime and compute the difference and mean between the rf and rr
  # model predictions
  hamby224_lime_comparisons <- hamby224_test_explain %>%
    select(-data, -prediction) %>%
    group_by(training_data, case, bin_continuous, quantile_bins, nbins, use_density) %>%
    slice(1) %>%
    ungroup() %>%
    select(training_data, bin_continuous, quantile_bins, nbins, use_density, 
           set, case, rf_features, rfscore, model_prediction, model_r2) %>%
    mutate(diff = rfscore - model_prediction,
           mean = (rfscore + model_prediction) / 2)

@@

```{r}
hamby224_lime_results <- hamby224_lime_comparisons %>%
  filter(!is.na(rfscore)) %>%
  group_by(training_data, bin_continuous, quantile_bins, nbins, use_density, set) %>%
  summarise(mse = (sum(diff)^2) / length(diff),
            ave_r2 = mean(model_r2)) %>%
  arrange(training_data, set)
```

```{r}
hamby224_lime_results %>%
  group_by(training_data, set) %>%
  arrange(mse) %>%
  slice(1:2)
```

```{r}
hamby224_lime_results %>%
  ungroup() %>%
  mutate(bc_qb = as.character(factor(bin_continuous):factor(quantile_bins)),
         nbins = factor(ifelse(bin_continuous == FALSE, NA, as.character(nbins)))) %>%
  ggplot(aes(x = nbins, y = mse, color = bc_qb, group = bc_qb)) +
  geom_point() + 
  geom_line() + 
  facet_grid(training_data ~ set) + 
  theme_bw()
```

```{r}
hamby224_lime_results %>%
  group_by(training_data, set) %>%
  arrange(desc(ave_r2)) %>%
  slice(1:2)
```

```{r}
hamby224_lime_results %>%
  ungroup() %>%
  mutate(bc_qb = as.character(factor(bin_continuous):factor(quantile_bins)),
         nbins = factor(ifelse(bin_continuous == FALSE, NA, as.character(nbins)))) %>%
  ggplot(aes(x = nbins, y = ave_r2, color = bc_qb, group = bc_qb)) +
  geom_point() + 
  geom_line() + 
  facet_grid(training_data ~ set) + 
  theme_bw()
```

@@

@@

```{r}
hamby224_test_explain %>%
  filter(training_data == "Full", set == "Set 1", nbins == "2") %>%
  ggplot(aes(x = feature_desc)) + 
  geom_bar() +
  coord_flip() + 
  facet_grid(set ~ .)
```

```{r}
hamby224_test_explain %>%
  filter(training_data == "Full", set == "Set 1", nbins == "2") %>%
  mutate(rfscore = round(rfscore, 3)) %>%
  select(case, bullet1, bullet2, land1, land2, rfscore) %>%
  distinct() %>%
  bind_rows(hamby224_test_explain %>%
                  filter(is.na(rfscore), set == "Set 1") %>%
                  select(case, bullet1, bullet2, land1, land2, rfscore)) %>%
  ggplot(aes(x = land1, y = land2, label = bullet1, label2 = bullet2,
             text = paste('Bullets Compared: ', bullet1, "-", land1, 
                          "vs", bullet2, "-", land2,
                          '\nRandom Forest Score: ', 
                          ifelse(is.na(rfscore), "Missing due to tank rash", rfscore)))) +
  geom_tile(aes(fill = rfscore)) +
  facet_grid(bullet2 ~ bullet1, scales = "free") +
  theme_minimal() +
  scale_fill_gradient2(low = "darkgrey", high = "darkorange", midpoint = 0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  labs(x = "", y = "", fill = "RF Score")
```

>>>>

# Load libraries
library(tidyverse)
library(bulletr)
library(randomForest)
library(lime)
library(cowplot)
library(purrr)
library(future)
library(furrr)

@@

@@

```{r fig.height = 5, fig.width = 8}
# Create plots of the feature distributions colored by samesource for the training data
hamby173and252_train %>% 
  select(rf_features, samesource) %>%
  gather(key = feature, value = value, 1:9) %>%
  select(feature, value, samesource) %>%
  ggplot(aes(x = value, fill = samesource)) + 
  geom_histogram(bins = 30) + 
  facet_wrap( ~ feature, scales = "free") +
  labs(x = "Variable Value", y = "Frequency", fill = "Same Source?") +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r fig.height = 5, fig.width = 8}
# Create plots of the feature distributions colored by samesource for the training data
# using the position = "fill" option
hamby173and252_train %>% 
  select(rf_features, samesource) %>%
  gather(key = feature, value = value, 1:9) %>%
  select(feature, value, samesource) %>%
  ggplot(aes(x = value, fill = samesource)) + 
  geom_histogram(position = "fill", bins = 30) + 
  facet_wrap( ~ feature, scales = "free") +
  labs(x = "Variable Value", y = "Proportion", fill = "Same Source?") +
  theme_bw() +
  theme(legend.position = "bottom")
```

@@

  # Specify the options to use with lime
  hamby224_lime_inputs <- data.frame(bin_continuous = c(rep(TRUE, 10), rep(FALSE, 2)),
                                     quantile_bins = c(rep(TRUE, 5), rep(FALSE, 5), 
                                                       rep(TRUE, 2)),
                                     nbins = c(rep(2:6, 2), rep(4, 2)),
                                     use_density = c(rep(TRUE, 10), TRUE, FALSE)) %>%
    mutate(case = 1:length(bin_continuous)) %>%
    select(case, bin_continuous, nbins, quantile_bins, use_density)

  # Tell R to run the upcoming code in parallel
  plan(multiprocess)

  # Apply lime to the full training data with the specified input options
  hamby224_lime_explain <- future_pmap(.l = as.list(hamby224_lime_inputs %>% 
                                                    select(-case)),
             .f = run_lime, # run_lime is one of my helper functions
             train = hamby173and252_train %>% select(rf_features),
             test = hamby224_test %>% arrange(case) %>% select(rf_features) %>% na.omit(),
             rfmodel = as_classifier(rtrees),
             label = "TRUE",
             nfeatures = 3,
             seed = TRUE)

  # Separate the lime and explain function results from the full data
  hamby224_lime <- map(hamby224_lime_explain, function(list) list$lime)
  hamby224_explain <- map_df(hamby224_lime_explain, function(list) list$explain)

  # Name the items in the lime list
  names(hamby224_lime) <- map_chr(1:dim(hamby224_lime_inputs)[1], function(case) 
      sprintf("case: bin_continuous = %s, quantile_bins = %s, nbins = %0.f, use_density = %s",
              hamby224_lime_inputs$bin_continuous[case],
              hamby224_lime_inputs$quantile_bins[case],
              hamby224_lime_inputs$nbins[case],
              hamby224_lime_inputs$use_density[case]))

@@

  # Join the data and the explanations and edit and add additional variables
  # Create the feature bin labels using my function "bin_labeller"
  hamby224_test_explain <- hamby224_test %>%
    mutate(case = as.character(case)) %>%
    full_join(hamby224_explain, by = "case") %>%
    mutate(case = factor(case),
           feature_desc = factor(feature_desc),
           feature_bin = pmap_chr(list(feature = feature, 
                                  feature_value = feature_value,
                                  b_c = bin_continuous,
                                  q_b = quantile_bins,
                                  n_b = nbins,
                                  u_d = use_density),
                            .f = bin_labeller, # bin_labeller is one of my helper functions
                            bin_data = hamby224_bin_boundaries,
                            case_info = hamby224_lime_inputs)) %>%
    mutate(feature = factor(feature),
           nbins = factor(nbins),
           feature_number = readr::parse_number(feature_desc),
           strictly_less = FALSE) %>%
    arrange(nbins)

@@

  # Reorder the variables of feature_desc and feature_bin for plotting purposes
  hamby224_test_explain <- hamby224_test_explain %>%
    mutate(feature_desc = reorder(feature_desc, strictly_less),
           feature_desc = reorder(feature_desc, feature_number),
           feature_desc = reorder(feature_desc, as.numeric(feature))) %>%
    select(bin_continuous:use_density, case:feature_desc, feature_bin:strictly_less, 
           data, prediction)

@@

  # Create a data frame with the interesting information relating to the differe
  # evaluations of lime and compute the difference and mean between the rf and rr
  # model predictions
  hamby224_lime_comparisons <- hamby224_test_explain %>%
    select(-data, -prediction) %>%
    group_by(case, bin_continuous, quantile_bins, nbins, use_density) %>%
    slice(1) %>%
    ungroup() %>%
    select(bin_continuous, quantile_bins, nbins, use_density, 
           set, case, rf_features, rfscore, model_prediction, model_r2) %>%
    mutate(diff = rfscore - model_prediction,
           mean = (rfscore + model_prediction) / 2)

@@

```{r}
hamby224_lime_results <- hamby224_lime_comparisons %>%
  filter(!is.na(rfscore)) %>%
  group_by(bin_continuous, quantile_bins, nbins, use_density, set) %>%
  summarise(mse = (sum(diff)^2) / length(diff),
            ave_r2 = mean(model_r2)) %>%
  arrange(set)
```

```{r}
hamby224_lime_results %>%
  arrange(mse)
```

```{r}
hamby224_lime_results %>%
  ungroup() %>%
  mutate(bc_qb = as.character(factor(bin_continuous):factor(quantile_bins)),
         nbins = factor(ifelse(bin_continuous == FALSE, NA, as.character(nbins)))) %>%
  ggplot(aes(x = nbins, y = mse, color = bc_qb, group = bc_qb)) +
  geom_point() + 
  geom_line() + 
  facet_grid( ~ set) + 
  theme_bw()
```

```{r}
hamby224_lime_results %>%
  group_by(set) %>%
  arrange(desc(ave_r2)) %>%
  slice(1:2)
```

```{r}
hamby224_lime_results %>%
  ungroup() %>%
  mutate(bc_qb = as.character(factor(bin_continuous):factor(quantile_bins)),
         nbins = factor(ifelse(bin_continuous == FALSE, NA, as.character(nbins)))) %>%
  ggplot(aes(x = nbins, y = ave_r2, color = bc_qb, group = bc_qb)) +
  geom_point() + 
  geom_line() + 
  facet_grid(~ set) + 
  theme_bw()
```

@@

## Assessing the Variability of the LIME Results

```{r}
# Perform the sensitivity analysis if not already saved
if(!file.exists("../data/hamby224_sensitivity_inputs.rds")) {

  # Specify the inputs for the sensitivity analysis 
  sensitivity_inputs <- data.frame(bin_continuous = rep(TRUE, 100),
                                   quantile_bins = rep(c(TRUE, FALSE), each = 50),
                                   nbins = rep(2:6, each = 10),
                                   use_density = rep(TRUE, 100)) %>%
    mutate(case = 1:length(bin_continuous)) %>%
    select(case, bin_continuous, nbins, quantile_bins, use_density)

  # Tell R to run the upcoming code in parallel
  plan(multiprocess)

  # Run lime for the sensitivity analysis
  sensitivity_outputs <- future_pmap(.l = as.list(sensitivity_inputs %>% select(-case)),
             .f = run_lime, # run_lime is one of my helper functions
             train = hamby173and252_train %>% select(rf_features),
             test = hamby224_test %>% arrange(case) %>% select(rf_features) %>% na.omit(),
             rfmodel = as_classifier(rtrees),
             label = "TRUE",
             nfeatures = 3,
             seed = FALSE) %>%
    map_df(function(list) list$explain) %>%
    mutate(rep = factor(rep(rep(1:10, each = dim(hamby224_test %>% na.omit())[1] * 3), 10))) %>%
    full_join(hamby224_test %>% na.omit() %>% 
                mutate(case = as.character(case)), by = "case") %>%
    mutate(case = factor(case)) %>%
    select(case, model_r2:feature_weight, bin_continuous:rep, set:land2, rfscore, samesource)

  # Save the sensitivity inputs and outputs
  saveRDS(sensitivity_inputs, "../data/hamby224_sensitivity_inputs.rds")
  saveRDS(sensitivity_outputs, "../data/hamby224_sensitivity_outputs.rds")

  # Load in the sensitivity inputs and outputs
  sensitivity_inputs <- readRDS("../data/hamby224_sensitivity_inputs.rds")
  sensitivity_outputs <- readRDS("../data/hamby224_sensitivity_outputs.rds")

}

sensitivity_results <- sensitivity_outputs %>%
  mutate(diff = rfscore - model_prediction) %>%
  group_by(rep, bin_continuous, nbins, quantile_bins, use_density) %>%
  summarise(mse = sum((diff)^2) / length(diff))

sensitivity_results %>%
  group_by(bin_continuous, nbins, quantile_bins, use_density) %>%
  summarise(mean = mean(mse),
            sd = sd(mse))

sensitivity_results %>%
  ggplot(aes(x = factor(nbins), y = mse)) + 
  geom_boxplot() + 
  geom_point()
```

@@

```{r}
hamby224_test_explain %>%
  filter(set == "Set 1", nbins == "2") %>%
  ggplot(aes(x = feature_desc)) + 
  geom_bar() +
  coord_flip() + 
  facet_grid(set ~ .)
```

```{r}
hamby224_test %>%
  filter(set == "Set 1") %>%
  ggplot(aes(x = land1, y = land2, label = bullet1, label2 = bullet2,
             text = paste('Bullets Compared: ', bullet1, "-", land1, 
                          "vs", bullet2, "-", land2,
                          '\nRandom Forest Score: ', 
                          ifelse(is.na(rfscore), "Missing due to tank rash", rfscore)))) +
  geom_tile(aes(fill = rfscore)) +
  facet_grid(bullet2 ~ bullet1, scales = "free") +
  theme_minimal() +
  scale_fill_gradient2(low = "darkgrey", high = "darkorange", midpoint = 0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  labs(x = "", y = "", fill = "RF Score")
```
